Seagrass mapping and monitoring \\ along the coasts of Crete, Greece

### 2011-02-16 MScDiplom_Lemenkova.tex
finished references of Figures (links to text); updated Tables

\documentclass[11pt]{article}
\special{papersize=140mm,210mm}
\usepackage{authblk}
\usepackage[english]{babel}
%\usepackage{authorindex}
\usepackage{color}
\usepackage[stable]{footmisc}
\usepackage{setspace}
\usepackage[printonlyused,withpage]{acronym}
\usepackage[left=30mm,right=20mm,top=30mm,bottom=35mm]{geometry}
\usepackage{pslatex}
%\usepackage{times}
%\usepackage{psnfss}
%\usepackage{glossary}
\usepackage{graphics}
\usepackage{graphicx}
\usepackage[numbers]{natbib}
\usepackage{calc}
\usepackage{amsmath}
\usepackage{longtable}
\usepackage{rotating}
\usepackage[svgnames,table]{xcolor}
\usepackage[subfigure]{tocloft}
%\usepackage[table]{xcolor}
\usepackage{paralist}
%\usepackage{makeidx}\makeindex
%\usepackage[makeindex,protected]{splitidx}
%\usepackage{index}
\usepackage{imakeidx}
\makeindex[name=concepts,title=Index of Concepts,intoc]
\makeindex[name=people,title=Index of People,intoc,columns=3,columnseprule]
\usepackage[fit]{truncate}
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhead[RO,LE]{\truncate{.95\headwidth}{\leftmark}}
\usepackage{fancyheadings}
\topmargin = 1pt
\headheight=10mm
\headsep=7pt
%\ensuremath
\usepackage{lastpage}
\usepackage{float}
\usepackage{subfig}
\usepackage{caption}
\usepackage{chngcntr}
\counterwithin{figure}{section}
\counterwithin{table}{section}
\usepackage{appendix}
\captionsetup{margin=10pt,font=small,labelfont=bf}
\usepackage{wrapfig}
\definecolor{darkblue}{rgb}{0.0,0.0,0.3}
\setcounter{tocdepth}{4}
\parindent 0pt
%\floatstyle{boxed} 
\restylefloat{figure}
\usepackage{hyperref}
\hypersetup{pdftitle={Seagrass mapping and monitoring along the coasts of Crete, Greece}, pdfauthor={Lemenkova Polina}, pdfsubject={MScThesis}, pdfcreator={Lemenkova Polina}, pdfproducer={Lemenkova Polina}, colorlinks=true,linkcolor=Navy, citecolor=Crimson, filecolor=magenta, urlcolor=blue,pdfkeywords={P.oceanica, seagrass, spectral reflectance, remote sensing, WASI, RTM, Google Earth, mapping, Crete}}

@@

\textbf{Thesis Assessment Board}:
\begin{itemize}
	\item Name Examiner 1 (Chair) \hfill{Prof. Dr. \textit{Andrew K. Skidmore}\index[people]{Skidmore A.K.} (ITC, Department of Natural Resource)}
	\item Name Examiner 2 \hfill{Ms. Ir. \textit{Louise M. van Leeuwen}\index[people]{Leeuwen, L. M. van} (ITC, Department of Natural Resource)}
	\item External Examiner \hfill{Dr. {\textit{Ma\l{}gorzata Roge-Wi\'sniewska}\index[people]{Roge-Wi\'sniewska, M.}} (University of Warsaw, Faculty of Geography and Regional Studies)}
\end{itemize}

\textbf{Member Supervisors}: \\
\begin{itemize}
	\item[* ] Mr. \textit{Valentijn Venus}\index[people]{Venus, V.}  \hfill{(University of Twente, ITC, Department of Natural Resource)}
	\item[* ] Dr. \textit{Albertus G. Toxopeus}\index[people]{Toxopeus, A.G.}  \hfill{(University of Twente, ITC, Department of Natural Resource)}
\end{itemize}

@@

@@

\subsection[\textit{In-situ} observations of the seagrass]{The \textit{in-situ} observations of the seagrass}
The traditional methods of \textit{in-situ} seagrass monitoring include in general the following standard
scheme \cite{McKenzieetal03}\label{McKenzieetal03}. The seagrass is being sampled on the selected sites using transect lines,
quadrant frame, single point markers, \ac{GPS}, laptops with installed software (\ac{GIS}), \ac{SCUBA} equipment, \ac{iPAQ} and other auxiliary tools\label{page-15} (Fig.\ref{fig:3.3}). The seagrass sampling is taken on the regular way with observation points covering the study area with normal distribution.
During the measurement process, the vertical photograph of the measurements frame is taken, and the
following points are traditionally estimated: percentage of the seagrass cover within the quadrate,
species composition, sediment composition, canopy height, epyphyte abundance, algae percent cover,
count of microfauna and a specimen of seagrass is being taken. This scheme, well described by
McKenzie \cite{McKenzieetal03}\label{McKenzieetal03} is widely used and well-known among the marine biologists and
seagrass researchers. Applications of the \textit{in-situ} seagrass observations of the structuring epiphyte
community composition in the \textit{P.oceanica} ecosystems in Mediterranean Sea is, for example,
described by Villegas \cite{Villegas06}\label{Villegas06}. Realization of traditional methods for mapping seagrass usually
involves intensive and time-consuming \textit{in-situ} observations during the fieldwork, as, for example,
reported by Iverson and Bittaker \cite{Iverson86}\label{Iverson86}.  The results of the \textit{in-situ} measurements and observation are usually
managed and treated using integrated \ac{GIS} approach, as, e.g. reported by Schmieder \cite{Schmieder97}\label{Schmieder97}.\\
Other methods of seagrass \textit{in-situ} monitoring are based on the application of the active hydro-acoustic sonar sensors that send towards a sea floor a signal of energy and then collect the return echoes for the analysis (Fig.\ref{fig:sonar}).

@@

Transect method (Fig.\ref{fig:A.4}) is a common sampling technique in studies of the seagrass monitoring \cite{Shortis07}\label{Shortis07},
enabling the occurrences of the seagrass meadows to be recorded and counted in a systematic and
accurate way, detect spatial distribution of the single seagrass mattes to be properly identified,
without location bias.

@@

The videographic measurements and photos, enabling to detect types of the
seafloor, were captured by means of three underwater digital video cameras \ac{Olympus ST} 8000  (Fig.\ref{fig:3.3d})\label{page-26}, at
five minutes interval along the tracklog path (Fig.\ref{fig:3.6}).
research sampling design also included measurements of depths (Fig. \ref{fig:A.4}), because bathymetry is one of the
most determining factors for the seagrass locations. Measurements of depths were performed during
the fieldwork using cord, \ac{iPAQ}, and markers in order to assure that the videometric measurements
are taken at depths not more than four meters, in the shelf area.
As a result, nine transects were established of one m wide and 20 m long track, to cover seagrass beds
with videographic measurements. The \ac{GPS} allowed to capture measurement locations on the \ac{iPAQ},
encapsulated in a plastic waterproof Otterbox (Fig \ref{fig:3.3b}). The camera were adjusted horizontally by a leveler and
mounted under the bottom of the boat Fig.\ref{fig:A.51}\label{page-27} to capture video footage and imagery (Fig.\ref{fig:3.5}). The data
were taken at proper weather conditions: sunny, serene and cloud-free days with glassy sea state. 
The locations of the route were randomly selected in the areas of the Ligaria beach, to ensure most dense
coverage of the seagrass meadows in the research area. The underwater measurements of the seagrass
coverage were carried out by taking video footage and photos of ca 0.5 $m^2$ size each (Fig.\ref{fig:A.7}, Fig.\ref{fig:A.55}).
The results of the underwater videometric measurements include series of digital images helping to
classify seafloor cover types (Fig.\ref{fig:A.56}) and seagrass meadows, according to the differences in the structure,
colour, texture and shapes of the depicted objects\label{page-27}.\\
There are several types of the seagrass landscapes alon Ligaria beach, namely, seagrass meadows
continuously covering vast areas (Fig.\ref{fig:A.3} (a)), aggregated seagrass patches, represented by separate mattes with short irregular channels between them (Fig.\ref{fig:A.3} (b)) and isolated seagrass patches, or mattes, located separately from broad meadows (Fig.\ref{fig:A.3} (c), Fig.\ref{fig:A.56}).
The results of the underwater videometric measurements of the \ac{Olympus ST} cameras made during the
ship route include nine total tracklog routes in the selected research area, including series of
consequent images, completely covering the area under the boat path. The received data contain
information on seagrass presence within the study area, distribution of seagrass \textit{P.oceanica} meadows
and nature of the seafloor cover types: rocks, sandy, mixed (Fig.\ref{fig:A.6}, Fig.\ref{fig:A.5}, Fig.\ref{fig:A.3}). Seagrass species on the Ligaria and Agia Pelagia beach consist of \textit{P.oceanica}. The types of sediment on the Ligaria vary from coarse sand,
\textit{P.oceanica} patches, sand, mud, rock gravel and fine sand (Fig.\ref{fig:A.6}).

\subsection{Review of the collected data}
The remote sensing imagery was collected for the Crete island area covering the research period and
area, enabling the field data to be used for the calibration and validation. The available aerial and
satellite imagery are commonly used for the mapping of seagrass landscapes and their application is
proven by various research papers. The imagery includes satellite multi-spectral imagery (Landsat-TM,
ETM+) and aerial imagery from the Google Earth.
The overview of the collected data enables to summarize their following types:
\begin{itemize}
	\item[] 1. Optical spectra of \textit{P.oceanica}, carbonate sand, seawater with sediments and seawater measured in aquarium tank, without sediments, at different environmental conditions (e.g., Fig. \ref{fig:A.23})
	\item[]2. Aerial imagery from the \href{http://www.google.com/earth/index.html}{Google Earth}
	\item[]3. Satellite images from various open sources (e.g. Landsat, Tab \ref{tab:20}) (previews: Fig. \ref{fig:A.37})
	\item[]4. Results of underwater videometric measurements of the \ac{Olympus ST} cameras made during the ship route
\end{itemize}
The available satellite imagery (Fig.\ref{fig:A.38}) \label{page-30}and aerial images are read into the ArcGIS project.
The available broadband and hyperspectral remote sensing data are used for the mapping of the
seagrass in shelf areas no deeper than 4.0 meters, for the environmental monitoring in order to detect
the dynamics in changes of the seagrass distribution along Crete during the past 10 years using
different satellite images for 2000-2010.

\begin{figure}[h]
	\centering
	\includegraphics[scale=0.40]{UML_Dia.jpg}
	\caption{Work flow for the data acquisition, \href{http://live.gnome.org/Dia}{Dia} display}
	\label{fig:24}
\end{figure}

@@

\subsubsection{Implementation of statistical analysis}
The initial measured data were stored in raw-oriented format, so that re-formatting them into the
column-based layout was done using “transpose” command in Open Office or Excel (Fig. \ref{fig:A.11}). The next step included
calculation of the median, mean, quartiles and other statistical values at every data set (see \nameref{app},\label{dataprep} Fig. \ref{fig:A.25}, Tab. \ref{tab:17}).
After the preliminary analysis, the measured data were visualized using Gnuplot program, which
enables fine plotting of various datasets together: Fig \ref{fig:4.4}, Fig.\ref{fig:27}, Fig.\ref{fig:28}. The most acceptable method of interpolation was B\'ezier curve (Fig. \ref{fig:A.18}, Fig. \ref{fig:A.16}),
as it has trend-friendly graph better showing the general behavior of the curves at different wavebands
comparing to splines (see \nameref{app} for more results). 
Therefore, after several experiments with various interpolation techniques (Fig.\ref{fig:A.15}, Fig.\ref{fig:A.23}), we have chosen B\'ezier curves interpolation (Fig.\ref{fig:A.18}) which contains convex hulls made on its control points, and therefore is best suitable for our case: analysis of optical
properties of seawater.\\
To analyze average values of spectral reflectance of \textit{P.oceanica} by means of the measured sets we calculated 
mean values for the total set of measurements 1 – 350 (15th October): mean of min and max values, 
mean of average for each waveband, mean for statistical quartiles for each wavelength and mean for median, respectively. 
The adjusted averages evaluate mean values, needed for statistical analysis of the datasets. 
The outcome of this calculations is shown in the Tab.\ref{tab:5} and Tab.\ref{tab:6} (\nameref{app}). \\
As a result, we received mean values of spectral reflectance of \textit{P.oceanica} for extreme, average, median and both quartiles,
independent of the individual sets 1-350. \\
We used several statistical methods (Fig.\ref{fig:4.12})\label{stan} to asses data received as a result of measurements. 
The Student-t test\index[concepts]{Data!statistical analysis!Student-t testing},
one of the most commonly used techniques for testing a hypothesis on the basis
of a difference between sample means, is used for the data analysis. In our case the Student t-test demonstrates, if the variation
between two analyzed groups – values of spectral reflectance of seagrass \textit{P.oceanica} and sand - is
significant. Therefore, we use Student t-test to compare two sets of quantitative data of spectral
reflectance of \textit{P.oceanica} and sand, respectively, with samples collected independently of one
another. The Student-t test can be performed knowing just the means, standard deviation, and number
of data points. Therefore, we used the data (\nameref{app}) of means of spectral reflectance of the both
 cover types within data sets, their standard deviation, and the number of data points.
\pagebreak

@@

The \ac{LAD}\index[concepts]{Data!statistical analysis!LAD}, a popular optimization technique, were used to show the main trend of the distribution of spectral reflectance values along the spectra (Tab.\ref{tab:12}). Non-linear logistic analysis\index[concepts]{Data!statistical analysis!non-linear logistic analysis} is applied to specify and estimate a model of spectral distribution, in which the dependent variable (i.e. the value of spectral reflectance for each single observation) is not continuous, but discrete and independent for each case (Tab.\ref{tab:11}). \\

@@

As can be seen on Fig.\ref{fig:A.21}\label{A-23}, Fig.\ref{fig:A.22}\label{A-24} and Fig.\ref{fig:A.29}, it is a useful tool for finding repeating patterns for a single value of observations, such as the presence of a periodic signal which has been buried under noise. The autocorrelation analysis has been performed using both \ac{SPSS} and \ac{GRETL} software. 
The method of \textit{least squares}\index[concepts]{Data!statistical analysis!least squares, method of} minimizes the sum of the squares of the errors made in solving every single observation made. We tried different approaches of the \textit{least squares} methods: Tab.\ref{tab:15}, Tab.\ref{tab:18}, Tab.\ref{tab:27}.\\
To focus on the relationship between a dependent variable (each single observation of the spectral reflectance) and independent variables (three various depths: 0.5, 1.5 and 2.5 meters) we used the \textit{regression analysis}, which enables modeling and analyzing values of the spectral reflectance at several depths: Tab.\ref{tab:5}, Tab.\ref{tab:6}, Tab.\ref{tab:7}. 

@@

To estimate spectral density of a spectral reflectance measurements we used periodogram\index[concepts]{Data!statistical analysis!periodogram} function (Tab.\ref{tab:30}), which corresponds to the general spectrum of the observations with representation of a variable quantity.  
Thus, the periodogram (Fig.\ref{fig:A.31}) of the series of observation\index[concepts]{Data!statistical analysis!observations, series of} highlights so-called “dead spots” of low power between the frequencies we wish to exclude (can be seen clearly at the end of the graphs, e.g. Fig.\ref{fig:A.12} ) 

@@

areas of quartiles\index[concepts]{Data!statistical analysis!quartiles Q1-Q3} (Q1, Q3, shaded areas on graph, Fig. \ref{fig:4.10}, Fig.\ref{fig:4.14}), which cover most probable data distribution (spectral reflectance) for each spectral band. In x-axis displayed are the areas of 400 – 950 nm in

@@

\begin{figure}
	\begin{center}
		\includegraphics[scale=0.27]{Fig-37.jpg}
		\caption{Google aerial images incorporated into the GIS project: fragment of \href{http://www.esri.com/software/arcgis/index.html}{ArcGIS} layout}
		\label{fig:4.17}
	\end{center}
\end{figure}

>>>>

\documentclass[11pt]{article}
\special{papersize=140mm,210mm}
\usepackage{authblk}
\usepackage[english]{babel}
%\usepackage{authorindex}
\usepackage{color}
\usepackage[stable]{footmisc}
\usepackage{setspace}
\usepackage[printonlyused,withpage]{acronym}
\usepackage[left=30mm,right=20mm,top=30mm,bottom=35mm]{geometry}
\usepackage{pslatex}
%\usepackage{times}
%\usepackage{psnfss}
%\usepackage{glossary}
\usepackage{graphics}
\usepackage{graphicx}
\usepackage[numbers]{natbib}
\usepackage{calc}
\usepackage{amsmath}
\usepackage{longtable}
\usepackage{rotating}
\usepackage[svgnames,table]{xcolor}
\usepackage[subfigure]{tocloft}
%\usepackage[table]{xcolor}
\usepackage{paralist}
%\usepackage{makeidx}\makeindex
%\usepackage[makeindex,protected]{splitidx}
%\usepackage{index}
\usepackage{imakeidx}
\makeindex[name=concepts,title=Index of Concepts,intoc]
\makeindex[name=people,title=Index of People,intoc,columns=3,columnseprule]
\usepackage[fit]{truncate}
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhead[RO,LE]{\truncate{.95\headwidth}{\leftmark}}
\usepackage{fancyheadings}
\topmargin = 1pt
\headheight=10mm
\headsep=7pt
%\ensuremath
\usepackage{lastpage}
\usepackage{float}
\usepackage{subfig}
\usepackage{caption}
\usepackage{chngcntr}
\counterwithin{figure}{section}
\counterwithin{table}{section}
\usepackage{appendix}
\captionsetup{margin=10pt,font=small,labelfont=bf}
\usepackage{wrapfig}
\definecolor{darkblue}{rgb}{0.0,0.0,0.3}
\setcounter{tocdepth}{4}
\parindent 0pt
%\floatstyle{boxed} 
\restylefloat{figure}
\usepackage{hyperref}
\hypersetup{pdftitle={Seagrass mapping and monitoring along the coasts of Crete, Greece}, pdfauthor={Lemenkova Polina}, pdfsubject={MScThesis}, pdfcreator={Lemenkova Polina}, pdfproducer={Lemenkova Polina}, colorlinks=true,linkcolor=Navy, citecolor=Crimson, filecolor=magenta, urlcolor=blue,pdfkeywords={P.oceanica, seagrass, spectral reflectance, remote sensing, WASI, RTM, Google Earth, mapping, Crete}}
%\hypersetup{pdftitle={Seagrass mapping and monitoring along the coasts of Crete, Greece}, pdfauthor={Lemenkova Polina}, pdfsubject={MScThesis}, pdfcreator={Lemenkova Polina}, pdfproducer={Lemenkova Polina}, colorlinks=true,linkcolor=black, citecolor=black, filecolor=black, urlcolor=black,pdfkeywords={P.oceanica, seagrass, spectral reflectance, remote sensing, WASI, RTM, Google Earth, mapping, Crete}}

@@

\textbf{Thesis Assessment Board}:
\begin{itemize}
	\item Chair: Prof. Dr. \textit{Andrew K. Skidmore}\index[people]{Skidmore A.K.}
	\item External Examiner: Dr. {\textit{Ma\l{}gorzata Roge-Wi\'sniewska}\index[people]{Roge-Wi\'sniewska, M.}}
	\item Internal Examiner: Ms. Ir. \textit{Louise M. van Leeuwen}\index[people]{Leeuwen, L. M. van}
\end{itemize}

\textbf{Member Supervisors}: \\
\begin{itemize}
	\item[* ] First supervisor: Mr. \textit{Valentijn Venus}\index[people]{Venus, V.} 
	\item[* ] Second supervisor: Dr. \textit{Albertus G. Toxopeus}\index[people]{Toxopeus, A.G.}  
\end{itemize}

@@

	\acro{OLS}{\href{http://en.wikipedia.org/wiki/Ordinary_least_squares}{Ordinary Least Squares}}

@@

\subsection[\textit{In-situ} observations of the seagrass]{The \textit{in-situ} observations of the seagrass}
The traditional methods of \textit{in-situ} seagrass monitoring include in general the following standard
scheme \cite{McKenzieetal03}\label{McKenzieetal03}. The seagrass is being sampled on the selected sites using transect lines,
quadrant frame, single point markers, \ac{GPS}, laptops with installed software (\ac{GIS}), \ac{SCUBA} equipment, \ac{iPAQ} and other auxiliary tools\label{page-15} (Fig.\ref{fig:3.3}). The seagrass sampling is taken on the regular way with observation points covering the study area with normal distribution (Fig.\ref{fig:A.20}, Fig.\ref{fig:A.19}\label{page-norm}).
During the measurement process, the vertical photograph of the measurements frame is taken, and the
following points are traditionally estimated: percentage of the seagrass cover within the quadrate,
species composition, sediment composition, canopy height, epyphyte abundance, algae percent cover,
count of microfauna and a specimen of seagrass is being taken. This scheme, well described by
McKenzie \cite{McKenzieetal03}\label{McKenzieetal03} is widely used and well-known among the marine biologists and
seagrass researchers. Applications of the \textit{in-situ} seagrass observations of the structuring epiphyte
community composition in the \textit{P.oceanica} ecosystems in Mediterranean Sea is, for example,
described by Villegas \cite{Villegas06}\label{Villegas06}. Realization of traditional methods for mapping seagrass usually
involves intensive and time-consuming \textit{in-situ} observations during the fieldwork, as, for example,
reported by Iverson and Bittaker \cite{Iverson86}\label{Iverson86}.  The results of the \textit{in-situ} measurements and observation are usually
managed and treated using integrated \ac{GIS} approach, as, e.g. reported by Schmieder \cite{Schmieder97}\label{Schmieder97}.\\
Other methods of seagrass \textit{in-situ} monitoring are based on the application of the active hydro-acoustic sonar sensors that send towards a sea floor a signal of energy and then collect the return echoes for the analysis (Fig.\ref{fig:sonar}).

@@

Transect method (Fig.\ref{fig:A.4}) is a common sampling technique in studies of the seagrass monitoring \cite{Shortis07}\label{Shortis07},
enabling the occurrences of the seagrass meadows to be recorded and counted in a systematic and
accurate way, detect spatial distribution of the single seagrass mattes to be properly identified,
without location bias (Fig.\ref{fig:A.7})\label{page-A12}.

@@

The videographic measurements and photos, enabling to detect types of the
seafloor, were captured by means of three underwater digital video cameras \ac{Olympus ST} 8000  (Fig.\ref{fig:3.3d})\label{page-26}, at
five minutes interval along the tracklog path (Fig.\ref{fig:3.6}).
research sampling design also included measurements of depths (Fig. \ref{fig:A.4}), because bathymetry is one of the
most determining factors for the seagrass locations. Measurements of depths were performed during
the fieldwork using cord, \ac{iPAQ}, and markers (Fig.\ref{fig:A.55})\label{page-A55} in order to assure that the videometric measurements
are taken at depths not more than four meters, in the shelf area.
As a result, nine transects were established of one m wide and 20 m long track, to cover seagrass beds
with videographic measurements. The \ac{GPS} allowed to capture measurement locations on the \ac{iPAQ},
encapsulated in a plastic waterproof Otterbox (Fig \ref{fig:3.3b}). The camera were adjusted horizontally by a leveler and
mounted under the bottom of the boat Fig.\ref{fig:A.51}\label{page-27} to capture video footage and imagery (Fig.\ref{fig:3.5}). The data
were taken at proper weather conditions: sunny, serene and cloud-free days with glassy sea state. 
The locations of the route were randomly selected in the areas of the Ligaria beach, to ensure most dense
coverage of the seagrass meadows in the research area. The underwater measurements of the seagrass
coverage were carried out by taking video footage and photos of ca 0.5 $m^2$ size each (Fig.\ref{fig:A.7}, Fig.\ref{fig:A.55}).
The results of the underwater videometric measurements include series of digital images helping to
classify seafloor cover types (Fig.\ref{fig:A.56}) and seagrass meadows, according to the differences in the structure,
colour, texture and shapes of the depicted objects\label{page-27}.\\
There are several types of the seagrass landscapes alon Ligaria beach, namely, seagrass meadows
continuously covering vast areas (Fig.\ref{fig:A.3} (a)), aggregated seagrass patches, represented by separate mattes with short irregular channels between them (Fig.\ref{fig:A.3} (b)) and isolated seagrass patches, or mattes, located separately from broad meadows (Fig.\ref{fig:A.3} (c), Fig.\ref{fig:A.56}).
The results of the underwater videometric measurements of the \ac{Olympus ST} cameras made during the
ship route include nine total tracklog routes in the selected research area, including series of
consequent images, completely covering the area under the boat path. The received data contain
information on seagrass presence within the study area, distribution of seagrass \textit{P.oceanica} meadows
and nature of the seafloor cover types: rocks, sandy, mixed (Fig.\ref{fig:A.6}, Fig.\ref{fig:A.5}, Fig.\ref{fig:A.3}). Seagrass species on the Ligaria and Agia Pelagia beach consist of \textit{P.oceanica}. The types of sediment on the Ligaria vary from coarse sand,
\textit{P.oceanica} patches, sand, mud, rock gravel and fine sand (Fig.\ref{fig:A.6}).

\subsection{Review of the collected data}
The remote sensing imagery was collected for the Crete island area covering the research period and
area, enabling the field data to be used for the calibration and validation. The available aerial and
satellite imagery are commonly used for the mapping of seagrass landscapes and their application is
proven by various research papers. The imagery includes satellite multi-spectral imagery (Landsat-TM,
ETM+) and aerial imagery from the Google Earth.
The overview of the collected data enables to summarize their following types:
\begin{itemize}
	\item[] 1. Optical spectra of \textit{P.oceanica}, carbonate sand, seawater with sediments and seawater measured in aquarium tank, without sediments, at different environmental conditions (e.g., Fig. \ref{fig:A.23})
	\item[]2. Aerial imagery from the \href{http://www.google.com/earth/index.html}{Google Earth}
	\item[]3. Satellite images from various open sources (e.g. Landsat, Tab \ref{tab:20}) (previews: Fig. \ref{fig:A.37})
	\item[]4. Results of underwater videometric measurements of the \ac{Olympus ST} cameras made during the ship route
\end{itemize}
The available satellite imagery (Fig.\ref{fig:A.38}, Tab.\ref{tab:34}) \label{page-30}and aerial images are read into the ArcGIS project.
The available broadband and hyperspectral remote sensing data are used for the mapping of the
seagrass in shelf areas no deeper than 4.0 meters (Fig.\ref{fig:A.39}, Fig.\ref{fig:A.40}, Fig.\ref{fig:A.41}, Fig.\ref{fig:A.42})\label{page-A45-47}, for the environmental monitoring in order to detect spatial distribution of the seagrass along Crete during the past ten years using
different satellite images for 2000-2010 (Fig.\ref{fig:24})\label{page-fig37}.

\begin{figure}[h]
	\centering
	\includegraphics[scale=0.40]{UML_Dia.jpg}
	\caption{Work flow for the data acquisition, \href{http://live.gnome.org/Dia}{Dia} display. Page: $<$\pageref{page-fig37}$>$}
	\label{fig:24}
\end{figure}

@@

\subsubsection{Implementation of statistical analysis}
The initial measured data were stored in raw-oriented format, so that re-formatting them into the
column-based layout was done using “transpose” command in Open Office or Excel (Fig. \ref{fig:A.11}). The next step included
calculation of the median, mean, quartiles and other statistical values at every data set (see \nameref{app},\label{dataprep} Fig. \ref{fig:A.25}, Tab. \ref{tab:17}).
After the preliminary analysis, the measured data were visualized using Gnuplot program, which
enables fine plotting of various datasets together: Fig \ref{fig:4.4}, Fig.\ref{fig:27}, Fig.\ref{fig:28}. The most acceptable method of interpolation was B\'ezier curve (Fig. \ref{fig:A.18}, Fig. \ref{fig:A.16}),
as it has trend-friendly graph better showing the general behavior of the curves at different wavebands
comparing to splines (see \nameref{app} for more results). 
Therefore, after several experiments with various interpolation techniques (Fig.\ref{fig:A.15}, Fig.\ref{fig:A.23}), we have chosen B\'ezier curves interpolation (Fig.\ref{fig:A.18}) which contains convex hulls made on its control points, and therefore is best suitable for our case: analysis of optical
properties of seawater.\\
To analyze average values of spectral reflectance of \textit{P.oceanica} by means of the measured sets we calculated 
mean values for the total set of measurements 1 – 350 (15th October): mean of min and max values, 
mean of average for each waveband, mean for statistical quartiles for each wavelength and mean for median, respectively. 
The adjusted averages evaluate mean values of trend, needed for statistical analysis of the datasets (Fig.\ref{fig:A.32}). 
The outcome of this calculations is shown in the Tab.\ref{tab:5} and Tab.\ref{tab:6} (\nameref{app}). \\
As a result, we received mean values of spectral reflectance of \textit{P.oceanica} for extreme, average (Fig.\ref{fig:A.34}, Fig.\ref{fig:A.33})\label{page-ave}, median and both quartiles, independent of the individual sets 1-350. \\
We used several statistical methods (Fig.\ref{fig:4.12})\label{stan} to asses data received as a result of measurements. 
The Student-t test\index[concepts]{Data!statistical analysis!Student-t testing},
one of the most commonly used techniques for testing a hypothesis on the basis
of a difference between sample means, is used for the data analysis. In our case the Student t-test demonstrates, if the variation
between two analyzed groups – values of spectral reflectance of seagrass \textit{P.oceanica} and sand - is
significant. Therefore, we use Student t-test to compare two sets of quantitative data of spectral
reflectance of \textit{P.oceanica} and sand, respectively, with samples collected independently of one
another. The Student-t test can be performed knowing just the means, standard deviation, and number
of data points. Therefore, we used the data (\nameref{app}) of means of spectral reflectance of the both
 cover types within data sets, their standard deviation, and the number of data points.
\pagebreak

@@

The \ac{LAD}\index[concepts]{Data!statistical analysis!LAD}, a popular optimization technique, was used to show the main trend of the distribution of spectral reflectance values along the spectra (Tab.\ref{tab:12})\label{page-tab12}. Non-linear logistic analysis\index[concepts]{Data!statistical analysis!non-linear logistic analysis} is applied to specify and estimate a model of spectral distribution, in which the dependent variable (i.e. the value of spectral reflectance for each single observation) is not continuous, but discrete and independent for each case (Tab.\ref{tab:11}). \\

@@

As can be seen on Fig.\ref{fig:A.21}\label{A-23}, Fig.\ref{fig:A.22}\label{A-24} and Fig.\ref{fig:A.29}, it is a useful tool for finding repeating patterns for a single value of observations, such as the presence of a periodic signal which has been buried under noise. The autocorrelation analysis has been performed using both \ac{SPSS} and \ac{GRETL} software (Fig.\ref{tab:28}\label{page-a-c}). 
The method of \textit{least squares}\index[concepts]{Data!statistical analysis!least squares, method of} minimizes the sum of the squares of the errors made in solving every single observation made. We tried different approaches of the \textit{least squares} methods: Tab.\ref{tab:15}, Tab.\ref{tab:18}, Tab.\ref{tab:27}, \ac{OLS}: Fig.\ref{tab:15}\label{page-OLS}\\
To focus on the relationship between a dependent variable (each single observation of the spectral reflectance) and independent variables (three various depths: 0.5, 1.5 and 2.5 meters) we used the \textit{regression analysis}, which enables modeling and analyzing values of the spectral reflectance at several depths: Tab.\ref{tab:5}\label{page-reg}, Tab.\ref{tab:6}, Tab.\ref{tab:7}. 

@@

To estimate spectral density of a spectral reflectance measurements we used periodogram\index[concepts]{Data!statistical analysis!periodogram} function (Tab.\ref{tab:30})\label{page-period}, which corresponds to the general spectrum of the observations with representation of a variable quantity.  
Thus, the periodogram (Fig.\ref{fig:A.31})\label{page-A31} of the series of observation\index[concepts]{Data!statistical analysis!observations, series of} highlights so-called “dead spots” of low power between the frequencies we wish to exclude (can be seen clearly at the end of the graphs, e.g. Fig.\ref{fig:A.12} ) 

@@

areas of quartiles\index[concepts]{Data!statistical analysis!quartiles Q1-Q3} (Q1 - Q3, Fig.\ref{fig:A.30}\label{page-A30}, and shaded Q-Q areas on graph, Fig. \ref{fig:4.10}, Fig.\ref{fig:4.14}), which cover most probable data distribution (spectral reflectance) for each spectral band. In x-axis displayed are the areas of 400 – 950 nm in

@@

\begin{figure}
	\begin{center}
		\includegraphics[scale=0.27]{Fig-37.jpg}
		\caption{Google aerial images incorporated into the GIS project: fragment of \href{http://www.esri.com/software/arcgis/index.html}{ArcGIS} layout. Page: $<$\pageref{fig4-17}$>$}
		\label{fig:4.17}
	\end{center}
\end{figure}

It encompasses the integration of
high resolution aerial color Google Earth photography, spaceborne satellite imagery, assessment of
spectral signatures using \ac{WASI} software\index[concepts]{Seawater optical properties!RTM!WASI}, image processing by means of Erdas Imagine (Fig.\ref{fig:39}, Fig.\ref{fig:42}) and Arc \ac{GIS} based mapping. The use of \ac{GIS} for data incorporation (Fig.\ref{fig:4.17}), storage, analyses, visualizing and mapping enables to analyze environmental changes within seagrass landscapes based on data from various
sources: aerial and satellite images, geographically referenced maps of Crete island and results of
images classification showing areas of seagrass distribution.
The final mapping has been supported in ArcGIS through the data exporting, conversion and
integration of various data in one \ac{GIS}-project (Fig.\ref{fig:4.17})\label{fig4-17}. Data collected during the fieldwork, imagery
of the seagrass distribution are added into a \ac{GIS} dataset for the assessment and spatial analysis.
\paragraph{Accuracy assessment}\label{page-50acc}
We prepared the error confusion matrices (Fig.\ref{tab:kappa-1} and Fig.\ref{tab:kappa-2}) using $\kappa$ (Kappa) statistics to assess and evaluate accuracy of the classification. The accuracy assessment has been done using Erdas Imaging function Classifier/Accuracy Assessment. In the Accuracy Assessment viewer we have chosen utility Edit in order to generate random points throughout the classified image, and then chosen the Create or add random points dialog. After the points were generated, we entered the class values for the reference points from the supervised classification. In order to perform a proper accuracy assessment we needed about 300 points so that we defined the number of points for the selection. From the option parameters for points distribution we have chosen "stratified random" as better representing the variables. Then we evaluated the location of the points and determined their class value, which was done with ground truth points from Google Earth aerial photos for various data and seafloor imagery and with reference to the signature file, to find the numeric value that has been assigned to each class previously. Finally, to print the errors matrix, we selected Accuracy Assessment viewer again and chosen there Report/ Options with turned on Error Matrix, Kappa Statistics and Accuracy Total. The error matrix is just comparing reference points of various seafloor cover types to the classified points (i.e. seagrass, carbonate sand, various land cover types on the coast). The Kappa coefficient takes into account chance agreement and thus, shows the reduction in errors generated by a classification process compared with the errors which could be received by a completely random classification – in other words, it evaluates quality of the classification. The results of the classification can be seen on the Fig. (see appendices). The overall map accuracy is 68\%, which means that 68\% of the pixels are classified to the correctly chosen seafloor cover type.
\pagebreak
